Portable Multi- and Many-Core Performance for Finite Difference Codes; Application to the Free-Surface Component of NEMO.

### 2017-06-22 docs/papers/nemolite2d_performance/nemolite2d_perf.tex
nemolite-paper: specify roofline is for Ivy Bridge and explain lack of SIMD performance benefit.

When using the OpenACC \texttt{kernels} directive the compiler creates
GPU kernels by parallelising only the outer loop of each loop nest.
For the majority of loop nests in NEMOLite2D all of the iterations are
independent and the loop bodies consist of just a handful of
executable statements. For small kernels the loop start-up cost can be
significant and therefore it is therefore beneficial to generate as
many threads as will fit on the GPU and then re-use them as
required. For this approach to be efficient we must expose the maximum
amount of parallelism to the GPU and we do this by instructing the
compiler to parallelise both levels (i.e. \texttt{ji} and \texttt{jj})
of a loop nest. This is achieved using the OpenACC \texttt{loop
  collapse(2)} directive for all of the smaller kernels in NEMOLite2D.

@@

In contrast to all of the other NEMOLite2D kernels, the Momentum
kernels ($u$ and $v$) are relatively large in terms of the number of
executable statements they contain. In turn, this means that the
corresponding kernels will require more state and thus more registers
on the GPU. Therefore, rather than generating as many threads as will
fit on the GPU, it is more efficient only to generate as many as
required by the problem. Once slots become free on the GPU new threads
will launch. This minimises the overhead associated with thread
creation and eases register pressure.  The compiler can be persuaded
to adopt the above strategy through the use of the (PGI-specific)
\texttt{loop shortloop} directive which tells it that there is not
likely to be much thread re-use in the following loop. This directive
was applied to both the \texttt{ji} and \texttt{jj} loops in both of
the Momentum kernels.

@@

We have used different scales for the $y$-axes in each of the plots in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} in order to highlight the performance
differences between the code versions with a given compiler. However,
the best performance obtained on twelve threads (\textit{i.e.}  without
hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
Intel and Cray compilers, respectively. This serves to emphasise the
democratisation that the introduction of OpenMP has had; for the
serial case the Cray- and Intel-compiled exectables were a factor of
two faster than the Gnu-compiled binary
(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
Gnu-compiled binary is only 6\% slower than that produced by the Intel
compiler and is 13\% \textit{faster} that that of the Cray compiler. In
part this situation comes about because of the effect that adding the
OpenMP compiler flag has on the optimisations performed by the
compiler.  In particular, the Cray compiler no longer vectorises the
key Momentum section, despite the directive added during the serial
optimisation work. This defficiency has been reported to Cray.

@@

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
use the Roofline Model~\citep{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and use the
STREAM~\citep{stream} and LINPACK~\citep{linpack} benchmarks in order to
get the upper bounds on the memory bandwidth and floating-point
operations per second (FLOPS) for the test hardware. Since we are
using an Intel Ivybridge CPU we used the Intel Math Kernel Library
implementation of LINPACK.

A key component of the Roofline model is the Operational or Arithmetic
Intensity of the code being executed:
\[
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\]
We calculated this quantity manually by examining the
source code and counting the number of memory references and
arithmetic operations that it contained. In doing this counting we
assume that any references to adjacent array elements (\textit{e.g.}
$u(i,j)$ and $u(i+1,j)$) are fetched in a single cache line and thus
only count once.

@@

In order to aid our understanding of kernel performance we have
developed a tool capable of parsing Fortran code and generating a
Directed Acyclic Graph (DAG) of the data flow. This tool eases the
laborious and error-prone process of counting memory accesses and
FLOPs as well as providing information on those operations that are
rate-limiting or on the critical path. Using the details of the Intel
Ivy Bridge microarchitecture published by Fog~\citep{fog_microarch,
  fog_instructions} we have constructed performance estimates of the
NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
(ILP), \textit{i.e.} assuming that all nodes in the DAG are executed
in serial, we get a lower-bound performance estimate of $0.6391 \times
CLOCK\_SPEED$ FLOPS which gives 2.46~GFLOPS at a clock speed of
3.85~GHz.

@@

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
our performance estimate. This is because we have incorporated the
effect of SSE into that estimate by simply assuming perfect
vectorisation which gives a performance increase of a factor of two.

>>>>

When using the OpenACC \texttt{kernels} directive the compiler creates
GPU kernels by parallelising only the outer loop of each loop nest.
For the majority of loop nests in NEMOLite2D all of the iterations are
independent and the loop bodies consist of just a handful of
executable statements. For small kernels the loop start-up cost can be
significant and therefore it is beneficial to generate as many threads
as will fit on the GPU and then re-use them as required (\textit{i.e.} if the
data size is greater than the number of threads). For this approach to
be efficient we must expose the maximum amount of parallelism to the
GPU and we do this by instructing the compiler to parallelise both
levels (i.e. \texttt{ji} and \texttt{jj}) of a loop nest. This is
achieved using the OpenACC \texttt{loop collapse(2)} directive for all
of the smaller kernels in NEMOLite2D.

@@

In contrast to all of the other NEMOLite2D kernels, the Momentum
kernels ($u$ and $v$) are relatively large in terms of the number of
executable statements they contain. In turn, this means that the
corresponding kernels will require more state and thus more registers
on the GPU. Therefore, rather than generating as many threads as will
fit on the GPU, it is more efficient only to generate as many as
required by the problem so as to minimise register pressure. Once
slots become free on the GPU, new threads will launch with the
associated cost amortized by the longer execution time of the larger
kernel. The compiler can be persuaded
to adopt the above strategy through the use of the (PGI-specific)
\texttt{loop shortloop} directive which tells it that there is not
likely to be much thread re-use in the following loop. This directive
was applied to both the \texttt{ji} and \texttt{jj} loops in both of
the Momentum kernels.

@@

We have used different scales for the $y$-axes in each of the plots in
Figures~\ref{FIG_gnu_omp_256},~\ref{FIG_intel_omp_256}
and~\ref{FIG_cray_omp_256} in order to highlight the performance
differences between the code versions with a given compiler. However,
the best performance obtained on twelve threads (\textit{i.e.}  without
hyperthreading) is 52.1, 55.6 and 46.3 (million points/s) for the Gnu,
Intel and Cray compilers, respectively. This serves to emphasise the
democratisation that the introduction of OpenMP has had; for the
serial case the Cray- and Intel-compiled exectables were a factor of
two faster than the Gnu-compiled binary
(Figure~\ref{FIG_opt_stages_256}). In the OpenMP version, the
Gnu-compiled binary is only 6\% slower than that produced by the Intel
compiler and is 13\% faster that that of the Cray compiler. In
part this situation comes about because of the effect that adding the
OpenMP compiler flag has on the optimisations performed by the
compiler.  In particular, the Cray compiler no longer vectorises the
key Momentum section, despite the directive added during the serial
optimisation work. This defficiency has been reported to Cray.

@@

Although we have investigated how the performance of the {PS}y{KA}l
version of NEMOLite2D compares with that of the original, we have not
addressed how efficient the original actually is. In order to do so we
consider the performance of the code on the Intel Ivy Bridge CPU and
use the Roofline Model~\citep{roofline} which provides a relatively
simple way of characterising the performance of a code in terms of
whether it is memory-bandwidth bound or compute bound. To do so we
follow the approach suggested in~\cite{para_pearls} and construct a
roofline model using the STREAM~\citep{stream} and
LINPACK~\citep{linpack} benchmarks in order to get the appropriate
upper bounds on the memory bandwidth and floating-point operations per
second (FLOPS). Since we are using an Intel Ivybridge CPU we used the
Intel Math Kernel Library implementation of LINPACK.

A key component of the Roofline model is the Operational or Arithmetic
Intensity (AI) of the code being executed:
\[
AI = \frac{\textrm{No. of floating-point operations}}{\textrm{Bytes fetched from memory}}
\]
We calculated this quantity manually by examining the
source code and counting the number of memory references and
arithmetic operations that it contained. In doing this counting we
assume that any references to adjacent array elements (\textit{e.g.}
$u(i,j)$ and $u(i+1,j)$) are fetched in a single cache line and thus
only count once.

@@

In order to aid our understanding of kernel performance we have
developed a tool, ``Habakkuk'' (https://github.com/arporter/habakkuk),
capable of parsing Fortran code and generating a Directed Acyclic
Graph (DAG) of the data flow. Habakkuk eases the laborious and
error-prone process of counting memory accesses and FLOPs as well as
providing information on those operations that are rate-limiting or on
the critical path. Using the details of the Intel Ivy Bridge
microarchitecture published by Fog~\citep{fog_microarch,
  fog_instructions} we have constructed performance estimates of the
NEMOLite2D kernel. By ignoring all Instruction-Level Parallelism
(ILP), \textit{i.e.} assuming that all nodes in the DAG are executed
in serial, we get a lower-bound performance estimate of $0.6391 \times
CLOCK\_SPEED$ FLOPS which gives 2.46~GFLOPS at a clock speed of
3.85~GHz.

@@

Enabling SIMD vectorisation for this kernel does not significantly
improve its performance (Figure~\ref{FIG_roofline}) and in fact,
limiting it to SSE instructions (vector width of two double-precision
floating point numbers) rather than AVX (vector width of four) was
found to produce a slightly more performant version. The poor
efficiency of the SIMD version is highlighted by the fact that the
corresponding kernel performance no longer falls within the bounds of
our performance estimate. This is because we have incorporated the
effect of SSE into that estimate by simply assuming perfect
vectorisation which gives a performance increase of a factor of two.
Further investigation of this issue revealed that several of the
NEMOLite2D kernels make frequent use of floating-point division
operations. Although SSE and AVX versions of the division operation
are available, on Ivy Bridge they do not provide any performance
benefit~\citep{fog_instructions}. A straightforward optimisation then
would be to alter the kernels in order to reduce the number of
division operations. However, kernel optimisation is outside the scope
of this paper since it breaks the PSyKAl separation of concerns.
